---
title: "FIN 452 QTP"
format: html
editor: visual
---

```{r, echo = FALSE, message = FALSE}
set.seed(123)
library(factoextra)
library(tidyquant)
library(tidyverse)
library(PerformanceAnalytics)
library(cluster)
library(gt)
```

```{r, echo = FALSE, message = FALSE}
# Function to calculate financial metrics for a given dataframe of stock prices
calculateStatistics <- function(price_data) {
  # Compute daily log returns
  returns_df <- price_data %>%
    group_by(symbol) %>%
    tq_transmute(select     = adjusted,
                 mutate_fun = periodReturn,
                 period     = "daily",
                 type       = "arithmetic",
                 col_rename = "daily_return") %>%
    ungroup()

  # Identify benchmark (SPY) returns for beta calculation
  market_returns <- returns_df %>%
    filter(symbol == "SPY") %>%
    select(date, market_return = daily_return)

  # Join market returns back to the main returns dataframe
  returns_with_market <- returns_df %>%
    left_join(market_returns, by = "date")

  # Compute metrics per ticker
  # Metrics include: average return, volatility, sharpe ratio, skewness, kurtosis, and asset beta
  summary_stats <- returns_with_market %>%
    filter(symbol != "SPY") %>% 
    group_by(symbol) %>%
    summarise(
      avg_return   = mean(daily_return, na.rm = TRUE),
      volatility   = sd(daily_return, na.rm = TRUE),
      sharpe_ratio = if(sd(daily_return, na.rm = TRUE) != 0) mean(daily_return, na.rm = TRUE) / sd(daily_return, na.rm = TRUE) else NA,
      skewness     = PerformanceAnalytics::skewness(daily_return),
      kurtosis     = PerformanceAnalytics::kurtosis(daily_return),
      asset_beta   = cov(daily_return, market_return, use = "complete.obs") / var(market_return, na.rm = TRUE)
    )

  return(summary_stats)
}
```

## Strategy Summary

My strategy revolves around using clusters in conjunction with a momentum based moving average model to identify and prioritize the top performing ETFs. Using the elbow test method I will identify the optimal number of clusters and compare the results of both the k means and hierarchical models. If they demonstrate a strong degree of clustering in the data I will then flesh out the portfolio adjustment process. Each cluster will be assigned a weight based on simulated model performance between 2000-01-01 and 2024-12-31. Each ETF within a particular cluster will be equally weighted with adjustments taking place on the first trading day of each month. Moving averages of denoised closing prices of 100 and 30 trading days will determine if each ETF will be held as a short or long position.

## Cluster Analysis

```{r, echo = FALSE, message = FALSE}
# Example usage with the tickers previously requested
tickers <- c("SPY", "XLB", "XLC", "XLE", "XLF", "XLI", "XLK", "XLP", "XLRE", "XLV", "XLU", "XLY")

# Pull historical data
historical_prices <- tickers %>%
  tq_get(get = "stock.prices",
         from = "2000-01-01",
         to = "2024-12-31") %>%
  group_by(symbol) %>%
  mutate(
    # Open-to-Close: Return within the trading day
    open_close_return = (close / open) - 1,
    # Close-to-Close: Total daily return
    close_close_return = close / lag(close) - 1,
    # Close-to-Open: Overnight return (Previous Close to Current Open)
    close_open_return = (open / lag(close)) - 1
  ) %>%
  ungroup()

# Results are stored in the ticker_metrics_df
ticker_metrics_df <- calculateStatistics(historical_prices)

scaled_metrics <- ticker_metrics_df %>% 
  select(avg_return, volatility, sharpe_ratio, asset_beta) %>% 
  scale()

max_k <- nrow(scaled_metrics) - 1

wss <- map_dbl(1:max_k, function(k){
  kmeans(scaled_metrics, centers = k, nstart = 25)$tot.withinss
})
tibble(k = 1:max_k, wss = wss) %>% 
  ggplot(aes(x = k, y = wss)) +
  geom_line(color = "blue", linewidth = 1) + geom_point(color = "blue") + scale_x_continuous(breaks = 1:max_k)+
  labs(title = "Optimizing the Number of Clusters",
       subtitle = "Metrics Tested: Average Return, Volatility, Sharpe Ratio, and Beta") + 
  xlab("Number of Clusters") + ylab("Within Cluster Sum of Squares")

```

Using the chart above we can narrow the the acceptable number of clusters to either 3 or 4. We will be proceeding with 3 clusters for further analysis as it will provide a clear delineation between high, mid, and lower growth ETFs.

```{r, echo = FALSE, message = FALSE}
#Compute Clusters Using K means
kmeans_res <- kmeans(scaled_metrics, centers = 3, nstart = 25)

ticker_metrics_df <- ticker_metrics_df %>% 
  mutate(cluster_kmeans = as.factor(kmeans_res$cluster))
```

```{r, echo = FALSE, message = FALSE}
#Compute clusters using heirarchical clustering
dist_mat <- dist(scaled_metrics, method = "euclidean")
cluster_result <- hclust(dist_mat, method = "ward.D2")
cluster_result$labels <- ticker_metrics_df$symbol

fviz_dend(cluster_result, k = 3, cex = 0.8, main = "Hierarchical Clustering of Assets", horiz = TRUE, rect = TRUE)
```

```{r, echo = FALSE, message = FALSE}
ticker_metrics_df <- ticker_metrics_df %>% 
  mutate(cluster_hclust = as.factor(cutree(cluster_result, k = 3)))

comparison_df <- ticker_metrics_df %>% 
  select(symbol, cluster_kmeans, cluster_hclust) %>% 
  arrange(cluster_kmeans)

comparison_df %>% gt() %>% gt::tab_header(title = "K Means Vs Heirarchical Clustering")




```

As demonstrated by the above chart K Means and Hierarchical Clustering are almost completely aligned with the only discrepancy being the placement of XLI. This indicates that a robust structure still exists between most of the assets but more analysis is required before settling on where to place XLI.

```{r, echo = FALSE, message = FALSE}
sil_kmean <- silhouette(kmeans_res$cluster, dist(scaled_metrics))

fviz_silhouette(sil_kmean) + labs(title = "K Means Cluster Quality")

sil_heir <- silhouette(cutree(cluster_result, k = 3), dist_mat)

fviz_silhouette(sil_heir) + labs(title = "Heirarchical Cluster Quality")

cluster_profiles <- ticker_metrics_df %>% 
  group_by(cluster_kmeans) %>% 
  summarise(
    avg_beta = mean(asset_beta),
    avg_return = mean(avg_return),
    avg_vol = mean(volatility),
    avg_sharpe = mean(sharpe_ratio),
    n_assets = n()
  ) %>% 
  arrange(desc(avg_beta))
```

We can see from the above silhouettes that hierarchical clustering has XLI as strongly negative, indicating it is likely misclassified. Therefore th clusters determined by the K Means method will be used for future analysis.

```{r, echo = FALSE, message = FALSE}
cluster_profiles %>% gt() %>% tab_header(title = "Cluster Performance Metrics")
#Assign descriptive names to clusters in comparison_df
comparison_df <- comparison_df %>%
mutate(cluster_name = case_match(cluster_kmeans,
"1" ~ "High Growth",
"2" ~ "Conservative",
"3" ~ "Market Performers",
.default = "Other"                              
))                                              
# Apply cluster descriptors to historical_prices
historical_prices <- historical_prices %>%
  left_join(comparison_df %>% select(symbol, cluster_name), by = "symbol")
```

In order to facilitate easier discussion about each cluster they will be given descriptive names based on the relative beta compared to SPY.

-   Cluster 1 = High Growth

-   Cluster 2 = Market Performers

-   Cluster 3 = Conservative

```{r, echo = FALSE, message = FALSE}
# Function to denoise prices using a Kalman Filter (Local Level Model)
kalman_denoise <- function(x, Q = 0.0001, R = 0.01) {
  # Q: Process noise (how much we expect the 'true' price to move)
  # R: Measurement noise (how much market jitter we want to filter out)

  n <- length(x)
  x_hat <- numeric(n) # Filtered state estimates
  P <- numeric(n)     # Error covariance

  # Initialization
  x_hat[1] <- x[1]
  P[1] <- 1.0

  for (t in 2:n) {
    # 1. Predict Step
    x_hat_minus <- x_hat[t-1]
    P_minus <- P[t-1] + Q

    # 2. Update Step (Correction)
    K <- P_minus / (P_minus + R) # Kalman Gain
    x_hat[t] <- x_hat_minus + K * (x[t] - x_hat_minus)
    P[t] <- (1 - K) * P_minus
  }
  return(x_hat)
}

# Apply the Kalman Filter to denoise each ticker's adjusted price
historical_prices <- historical_prices %>%
  group_by(symbol) %>%
  mutate(denoised_close = kalman_denoise(adjusted)) %>%
  ungroup()
```

```{r, echo = FALSE, message = FALSE}
# Remove SPY and create a new dataframe for portfolio analysis
historical_prices_no_spy <- historical_prices %>%
  filter(symbol != "SPY")

```

```{r, echo = FALSE, message = FALSE}
# Simple Moving Average Trading Strategy
# Par 1: 30-day SMA of denoised_close
# Par 2: 100-day SMA of denoised_close
# Signal: 1 if Par 1 > Par 2, -1 if Par 1 < Par 2

historical_prices_no_spy <- historical_prices_no_spy %>%
  group_by(symbol) %>%
  mutate(
    par1 = TTR::SMA(denoised_close, n = 30),
    par2 = TTR::SMA(denoised_close, n = 100),
    signal = case_when(
      par1 > par2 ~ 1,
      par1 < par2 ~ -1,
      .default = NA_real_
    ),
    # Current position: start with 1 (long) and follow signal when available
    pos = if_else(is.na(signal), 1, signal),
    # Returns of existing positions
    return_exist = if_else(pos == lag(pos, default = 1),close_close_return * pos, 0),
    # Return on portfolio changes (rebalancing at the open)
    new_returns = if_else(pos != lag(pos, default = 1), pos * open_close_return, 0),
    # Total return
    total_ret = return_exist + new_returns,
    # Cumulative returns
    cum_ret = cumprod(1 + coalesce(total_ret, 0))
  ) %>%
  group_by(symbol, month = floor_date(date, "month")) %>%
  mutate(cum_ret_monthly = cumprod(1 + coalesce(total_ret, 0))) %>%
  ungroup() %>%
  select(-month)
```

```{r, echo = FALSE, message = FALSE}
# Integrated Backtesting: Monthly Rebalancing with Daily SMA Long/Short Strategy
# This merges the signal logic into the portfolio value tracking

backtestStrategyPortfolio <- function(data, cluster_weights, initial_investment) {
  # 1. Identify rebalance dates (first trading day of each month)
  rebalance_dates <- data %>%
    mutate(month_year = floor_date(date, "month")) %>%
    group_by(month_year) %>%
    filter(date == min(date)) %>%
    ungroup() %>%
    pull(date) %>%
    unique() %>%
    sort()
  
  all_dates <- sort(unique(data$date))
  all_symbols <- unique(data$symbol)
  
  daily_portfolio <- data.frame(date = all_dates, total_value = NA_real_)
  
  # ticker_values: dollar value of each ticker's position contribution
  ticker_values <- setNames(rep(0, length(all_symbols)), all_symbols)
  # ticker_pos: direction (1 for long, -1 for short)
  ticker_pos <- setNames(rep(1, length(all_symbols)), all_symbols)
  
  current_total_val <- initial_investment
  
  for (i in seq_along(all_dates)) {
    today <- all_dates[i]
    today_data <- data[data$date == today, ]
    
    if (i == 1) {
      # Initialize at open of first day
      available <- today_data[!is.na(today_data$open) & !is.na(today_data$cluster_name) & !is.na(today_data$pos), ]
      counts <- table(available$cluster_name)
      if (nrow(available) > 0) {
        weights_vec <- cluster_weights[available$cluster_name]
        counts_vec <- as.numeric(counts[available$cluster_name])
        ticker_values[available$symbol] <- (initial_investment * weights_vec) / counts_vec
        ticker_pos[available$symbol] <- available$pos
      }
    } else {
      # A. Overnight performance (yesterday's positions)
      ret_co <- today_data$close_open_return[match(names(ticker_values), today_data$symbol)]
      ticker_values <- ticker_values * (1 + ticker_pos * coalesce(ret_co, 0))
      current_total_val <- sum(ticker_values)
      
      # B. Monthly Rebalance OR Daily Position Update
      if (today %in% rebalance_dates) {
        # Dynamically identify available tickers today
        available <- today_data[!is.na(today_data$open) & !is.na(today_data$cluster_name) & !is.na(today_data$pos), ]
        counts <- table(available$cluster_name)
        ticker_values[] <- 0 # Exit old positions
        if (nrow(available) > 0) {
          weights_vec <- cluster_weights[available$cluster_name]
          counts_vec <- as.numeric(counts[available$cluster_name])
          ticker_values[available$symbol] <- (current_total_val * weights_vec) / counts_vec
          ticker_pos[available$symbol] <- available$pos
        }
      } else {
        # Switch direction if signal changed
        new_pos <- today_data$pos[match(names(ticker_pos), today_data$symbol)]
        update_mask <- !is.na(new_pos) & ticker_values != 0
        ticker_pos[update_mask] <- new_pos[update_mask]
      }
    }
    
    # C. Intraday performance (today's updated positions)
    ret_oc <- today_data$open_close_return[match(names(ticker_values), today_data$symbol)]
    ticker_values <- ticker_values * (1 + ticker_pos * coalesce(ret_oc, 0))
    current_total_val <- sum(ticker_values)
    
    daily_portfolio$total_value[i] <- current_total_val
  }
  
  # Calculate Returns
  daily_portfolio <- daily_portfolio %>%
    mutate(
      daily_ret = total_value / lag(total_value) - 1,
      cum_ret = cumprod(1 + coalesce(daily_ret, 0))
    )
  
  return(daily_portfolio)
}

```

## Optimization

Now that clusters have been identified it is time to optimize the balance between them. As mentioned in the summary each ETF inside a cluster is equally weighted, for example if each cluster made up 33% of the portfolio, each of the four ETFs in the High Growth cluster would make up 25% of that 33% at every rebalancing day. Different portfolio weights will be simulated over the training period and then evaluated based on their Sharpe Ratio.

```{r, echo = FALSE, message = FALSE}
# Optimization: Grid Search for Optimal Cluster Weights
# We vary weights for the three clusters to maximize the Annualized Sharpe Ratio

# 1. Define the grid of weights (summing to 1)
weight_grid <- expand.grid(
  High_Growth = seq(0, 1, by = 0.2),
  Conservative = seq(0, 1, by = 0.2)
) %>%
  mutate(Market_Performers = 1 - (High_Growth + Conservative)) %>%
  filter(Market_Performers >= 0)

# 2. Iterate through the grid and calculate Sharpe Ratio
# Note: This might take a moment depending on the grid density
optimization_results <- weight_grid %>%
  rowwise() %>%
  mutate(
    sharpe_ratio = {
      w_vec <- c("High Growth" = High_Growth, 
                 "Conservative" = Conservative, 
                 "Market Performers" = Market_Performers)
      
      # Run backtest for these weights
      bt_res <- backtestStrategyPortfolio(historical_prices_no_spy, w_vec, 100000)
      
      # Calculate Annualized Sharpe Ratio
      # (Mean daily return / SD daily return) * sqrt(252)
      if (all(is.na(bt_res$daily_ret))) {
        NA_real_
      } else {
        mean(bt_res$daily_ret, na.rm = TRUE) / sd(bt_res$daily_ret, na.rm = TRUE) * sqrt(252)
      }
    }
  ) %>%
  ungroup()

# 3. Identify the best weight combination
best_weights <- optimization_results %>%
  arrange(desc(sharpe_ratio)) 
  best_weights %>% 
  gt() %>% 
  tab_header(title = "Optimal Cluster Weights")

# 4. Plot the optimization surface
ggplot(optimization_results, aes(x = High_Growth, y = Conservative, fill = sharpe_ratio)) +
  geom_tile() +
  scale_fill_viridis_c() +
  labs(title = "Cluster Weight Optimization Surface",
       subtitle = "Maximizing Annualized Sharpe Ratio",
       x = "Weight: High Growth",
       y = "Weight: Conservative",
       fill = "Sharpe Ratio") +
  theme_minimal()
```

From the above analysis we can see that based on Sharpe Ratio the optimal decision would be to focus solely the High Growth cluster. To ensure a more complete view of the models performance we will evaluate the cumulative returns of the top three Sharpe Ratios.

```{r, echo = FALSE, message = FALSE}
weights_opt <- c("High Growth" = as.numeric(best_weights[1,1]), "Conservative" = as.numeric(best_weights[1,2]), "Market Performers" = as.numeric(best_weights[1,3]))
opt_backtest <- backtestStrategyPortfolio(historical_prices_no_spy, weights_opt, 100000)

# Plot Cumulative Returns
ggplot(opt_backtest, aes(x = date, y = cum_ret)) +
  geom_line(color = "darkgreen") +
  theme_tq() +
  labs(title = "Optimal Sharpe Ratio Portfolio",
       subtitle = "High Growth = 1, Conservative = 0, Market Performers = 0",
       y = "Cumulative Return", x = "Date")

weights_2nd <- c("High Growth" = as.numeric(best_weights[2,1]), "Conservative" = as.numeric(best_weights[2,2]), "Market Performers" = as.numeric(best_weights[2,3]))
two_backtest <- backtestStrategyPortfolio(historical_prices_no_spy, weights_2nd, 100000)

ggplot(two_backtest, aes(x = date, y = cum_ret)) +
  geom_line(color = "orange") +
  theme_tq() +
  labs(title = "Second Best Sharpe Ratio Portfolio",
       subtitle = "High Growth = 0.8, Conservative = 0.2, Market Performers = 0",
       y = "Cumulative Return", x = "Date")

weights_3rd <- c("High Growth" = as.numeric(best_weights[3,1]), "Conservative" = as.numeric(best_weights[3,2]), "Market Performers" = as.numeric(best_weights[3,3]))
three_backtest <- backtestStrategyPortfolio(historical_prices_no_spy, weights_3rd, 100000)

ggplot(three_backtest, aes(x = date, y = cum_ret)) +
  geom_line(color = "purple") +
  theme_tq() +
  labs(title = "Third Best Sharpe Ratio Portfolio",
       subtitle = "High Growth = 0.8, Conservative = 0, Market Performers = 0.2",
       y = "Cumulative Return", x = "Date")
```

The solely High Growth strategy still performed the best in terms of overall returns; however the above charts also demonstrates that focusing solely on a handful of assets also carries its fair share of risk. There are several large and fast swings in the cumulative returns over the training set. Still mixing in either the Conservative or Market Performers clusters does not provide enough protection to make it worth giving up the extra returns. Therefore the High Growth Strategy will be used on test set.
